Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-04 application.tex
restructured
appendix.tex

\subsection{Radiative diffusion}
For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume  
temperature fluctuations are smoothed out by radiative diffusion. We
mode this in linear theory by writing the linearized RHS of
Eq. \ref{real_energy} as 
% we choose the heating term such that
% \begin{align}
%   Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
%     \left(T-T_{t=0}\right)\right], 
% \end{align} 
% where $k_\mathrm{rad}$ is the conduction coefficient to be defined
% implicitly. Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
%     c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
% \end{align}
% When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

@@

Now consider lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}. The curves shift inward, but are also
raised. In fact, lowering the opacity by an order of magnitude renders 
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This is
expected to stabilize the disk. Raising the opacity, as shown in 
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the fundamental mode is
restricted to larger radii.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
by choosing a sufficiently large $\khat$. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI,
should it develop in the inner disk, would only occur on very small
scales. However, our linear calculations suggest growth rates
eventually decay for large $\khat$.

Furthermore, modes with large $\khat$ may be subject to viscous decay
in a real disk. The viscous time for a perturbation lengthscale
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha_s\khat^2\Omega $, where
$\alpha_s$ is the standard alpha viscosity parameter
\citep{shakura73}. Setting $t_\mathrm{visc}$ to be longer than the
typical VSI growth timescale $1/\epsilon\Omega$ places an upper limit
$\khat \lesssim 30$ in a disk with $\alpha_s\sim 10^{-4}$ and
$\epsilon \sim 0.1$.

@@

\subsection{VSI in the MMSN}
Here we solve the linear problem in the MMSN with the scale and
height dependent thermal relaxation function defined by
Eq. \ref{beta_mmsn_simp}.  We use a vertical
domain $z_\mathrm{max}=5H$. In Fig. \ref{mmsn_overall} we plot growth
timescales $t_\mathrm{grow} \equiv 1/\nu$ associated with the
fundamental mode in units of the local orbital time
$P_\mathrm{orb}\equiv 2\pi/\Omega_k$.

In fact, for $1\lesssim \khat \lesssim 30$ we either only find the
fundamental mode, or that it is the most unstable. The associated 
growth rates are also insensitive to $z_\mathrm{max}$. However, for
large $\khat$ and/or $z_\mathrm{max}$, e.g. $\khat=50$ or 
$z_\mathrm{max}=10H$, we also found modes with shorter growth times 
than those shown in Fig. \ref{mmsn_overall}, but these disturbances are 
concentrated near the vertical boundaries
\citep{nelson13,mcnally14}. Such modes may not be robust because    
they are affected by boundary conditions. % , and high $\khat$ modes are
% subject to viscous decay in a real disk.

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see that growth timescales are $\sim
30$---40 orbits, and the VSI operates from a few AU to tens of
AU. Importantly, it is necessary to consider smaller scales as $r$
decreases as the disk generally becomes optically thick.

Fig. \ref{mmsn_overall} is roughly consistent with the 
mid-plane thermal relaxation timescales shown in
Fig. \ref{mmsn_bcrit_bcool}. For example, for $\khat=10$ we find 
$\beta \gtrsim \beta_\mathrm{crit}$ for $r\lesssim 10$AU at $z=0$; and
Fig. \ref{mmsn_overall} show stabilization inside $\sim 8$AU.

@@

\subsection{Comparing the VSI in the inner and outer parts of the MMSN} 
Here we compare the VSI in the MMSN at $r=50$AU and $r=5$AU. We first show in
Fig. \ref{beta_compare}  the thermal timescale $\beta$ for a
perturbation wavenumber $\khat=30$. For comparison we also plot
$\beta_\mathrm{crit}$ at these radii.

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with 
significantly reduced growth rates (when scaled by the local
rotation).  
%: Fig. \ref{mmsn_overall}) show that
%the growth time at

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer than the case with $\beta < \beta_\mathrm{crit}  \forall z$: we find $t_\mathrm{grow}\sim 40$ orbits at $r=50$AU,
but $t_\mathrm{grow}\sim 400$ orbits at $r=5$AU. Thus, introducing an
optically thick midplane reduces the efficiency of the VSI despite
$\beta < \beta_\mathrm{crit}$ away from the midplane.

>>>>

\subsection{Radiative diffusion}
For perturbations with lengthscales $l\gg
l_\mathrm{rad}$, we assume  
temperature fluctuations are smoothed out by radiative diffusion. We
model this in linear theory by writing the linearized RHS of
Eq. \ref{real_energy} as 
% we choose the heating term such that
% \begin{align}
%   Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
%     \left(T-T_{t=0}\right)\right], 
% \end{align} 
% where $k_\mathrm{rad}$ is the conduction coefficient to be defined
% implicitly. Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
%     c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
% \end{align}
% When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

@@

Now consider lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}. The curves shift inward, but are also
raised. In fact, lowering the opacity by an order of magnitude renders 
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This is
expected to stabilize the disk. Raising the opacity, as shown in 
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the instability is
restricted to larger radii.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
by choosing a sufficiently large $\khat$. We
illustrate this in Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at
each radius the value of $\khat$ such 
that $\beta = \beta_\mathrm{crit}$. We thus expect that the VSI,
should it develop in the inner disk, would only occur on very small
scales. However, our linear calculations suggest growth rates
eventually decay for large $\khat$.

Furthermore, modes with large $\khat$ may be subject to viscous decay
in a real disk. The viscous time for a perturbation lengthscale
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha_s\khat^2\Omega $, where
$\alpha_s$ is the standard alpha viscosity parameter
\citep{shakura73}. Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/\epsilon\Omega$ places an upper limit
$\khat^2\lesssim \epsilon/\alpha_s$. Inserting $\epsilon \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha_s\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha_s\sim
10^{-5}$. \citep[These low viscosity levels are required for the
VSI, see][]{nelson13}.

@@

\subsection{VSI in the MMSN}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation function defined by
Eq. \ref{beta_mmsn_simp}.  We use a vertical
domain $z_\mathrm{max}=5H$. In Fig. \ref{mmsn_overall} we plot growth
timescales $t_\mathrm{grow} \equiv 1/\nu$ associated with the
fundamental mode in units of the local orbital time
$P_\mathrm{orb}\equiv 2\pi/\Omega_k$.

In fact, for $1\lesssim \khat \lesssim 30$ we either only find the
fundamental mode, or that it is the most unstable. The associated 
growth rates are also insensitive to $z_\mathrm{max}$. However, for
large $\khat$ and/or $z_\mathrm{max}$, e.g. $\khat=50$ or 
$z_\mathrm{max}=10H$, we also found modes with shorter growth times 
than those shown in Fig. \ref{mmsn_overall}, but these disturbances are 
concentrated near the vertical boundaries
\citep{nelson13,mcnally14}. Such modes may not be robust because    
they are affected by boundary conditions and viscosity in a real
disk. % , and high $\khat$ modes are % subject to viscous decay in a real disk.

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits. 
Importantly, it is necessary to consider smaller scales as $r$
decreases as the disk generally becomes optically thick.

Fig. \ref{mmsn_overall} is roughly consistent with the 
mid-plane thermal relaxation timescales shown in
Fig. \ref{mmsn_bcrit_bcool}. For example, for $\khat=10$ we find 
$\beta \gtrsim \beta_\mathrm{crit}$ for $r\lesssim 10$AU at $z=0$; and
Fig. \ref{mmsn_overall} show stabilization inside $\sim 8$AU where
$t_\mathrm{grow}\to\infty$.

@@

\subsection{Comparing the VSI in the inner and outer parts of the MMSN} 
We compare the VSI in the MMSN at $r=50$AU and $r=5$AU. We first show in
Fig. \ref{beta_compare}  the thermal timescale $\beta$ for a
perturbation wavenumber $\khat=30$. For comparison we also plot
$\beta_\mathrm{crit}$ at these radii.

@@

In the outer disk ($r=50$AU) the disk is optically thin, $\beta$ 
does not depend on height and $\beta <
\beta_\mathrm{crit}$. Accordingly we expect, and find, the VSI to
operate at $r=50$AU. In the inner disk ($r=5$AU) the disk is optically
thick near the midplane and becomes optically thin away from it, such
that $\beta < \beta_\mathrm{crit}$ for $|z|\gtrsim1.25H$. In
fact, in this case  we also find the VSI to operate, but with much
reduced growth rates (when scaled by the local 
rotation).  
%: Fig. \ref{mmsn_overall}) show that
%the growth time at

The corresponding vertical velocity perturbations are shown in 
Fig. \ref{mmsn_eigenvz}. Although at $r=5$AU $\beta$ depends on $z$,
the disturbance still occupies the fluid column. This is probably due
to the vertically global nature of the fundamental mode. The fact that
we obtain the VSI despite $\beta > \beta_\mathrm{crit}$ near the
midplane suggests, but does not prove, that the VSI can operate
provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
are much longer at $r=5$AU than at $r=50$AU for which  $\beta <
\beta_\mathrm{crit}  \forall z$: we find $t_\mathrm{grow}\sim 40$
orbits at $r=50$AU, but $t_\mathrm{grow}\sim 400$ orbits at
$r=5$AU. Thus, introducing an optically thick midplane significantly
reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

### 2015-03-04 linear_adia.tex
restructured
appendix.tex

\subsection{Isothermal perturbations}\label{iso_pert}
Isothermal perturbations correspond to $\beta = 0$ or $\chi = 1$. This
case has been investigated numerically by \cite{nelson13} and 
\cite{mcnally14} in the thin disk limit, the latter of which employed
the VGSB framework.

In Appendix \S\ref{iso_discuss}, we give an analytical description of the
linear problem with $\beta=0$. There, we establish more general
results that do not require the thin disk approximation, namely 
a necessary condition for instability and an upper bound to the
instability growth rate. We also develop analytic solutions
with and without vertical shear in the thin disk limit, including
explicit expressions for wave frequencies and growth rates, the
methods and results for which motivates the solution procedure
employed in the following sections.

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in the case $\beta\neq 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For 
sufficiently small $\beta$, for which the perturbations may be 
considered nearly-isothermal, we expect the solution to only
differ slightly from a case with $\beta\equiv 0$.

@@

To determine the (assumed) small effect of introducing a short thermal
relaxation, we linearize Eq. \ref{nearly_iso_explicit} about the above
solution for $\beta\equiv0$. We write 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta,
\end{align}
with $\delta v_z$ and $\hat{\sigma}$ given by Eq. \ref{pert_basic}. We
may then seek  
\begin{align}
  \delta v_{z1} = a \zhat^2 + b,
\end{align}
where $a$, $b$ are constants.

@@

Introducing a finite thermal relaxation time introduces buoyancy
forces, which is stabilizing for sub-adiabatically stratified disks
($\gamma > 1$). The $\zhat^2$ dependence in $\delta v_{z1}$ then makes
sense because it is most significant for large $\zhat$, i.e. away from
the midplane where the effect of buoyancy first appears as $\beta$ is
increased from zero.

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. In Appendix \ref {iso_stable} we discuss
stable solutions without vertical shear which yield
$\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(Eq. \ref{iso_stable_disp}),  we are thus motivated to seek solutions
to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}\label{beta_crit0}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$.

From our discussion in \S\ref{relax_pert}, which indicate introducing
a finite thermal relaxation is stabilizing, we expect 
Eq. \ref{beta_crit0} is an upper limit for instability. For a thin
disk $\epsilon\ll 1$, we thus infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. That is, when
$\beta<\beta_\mathrm{crit}$, perturbations are  effectively isothermal. 
We will test Eq. \ref{iso_vsi_cond} numerically in
\S\ref{bcrit_num_test}.

>>>>

\subsection{Isothermal perturbations}\label{iso_pert}
Isothermal perturbations correspond to $\beta = 0$ or $\chi = 1$. This
case has been investigated numerically by \cite{nelson13} and 
\cite{mcnally14}, the latter of which employed the VGSB framework.

In Appendix \ref{iso_discuss}, we give an analytical description of the
linear problem with $\beta=0$. There, we establish more general
results that do not require the thin disk approximation, namely 
a necessary condition for instability (there must be vertical shear)
and an upper bound to the instability growth rate (it is limited by
the maximum vertical shear rate in the disk). We also develop analytic
solutions with and without vertical shear in the thin disk limit,
including explicit expressions for wave frequencies and growth rates. 
The methods used in Appendix \ref{iso_discuss} and results obtained
therein motivates the solution procedure employed in the following
sections.

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in the case $\beta\neq 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$t_c=\beta\Omega_k^{-1}$ from zero to a small but finite value? For 
sufficiently small $\beta$ % , for which the perturbations may be 
% considered nearly-isothermal,
we expect the solution to only differ slightly from a case with
$\beta\equiv 0$. We thus perturb a solution for $\beta=0$ to see the
effect of thermal relaxation.

@@

%To determine the (assumed) small effect of introducing a short thermal
%relaxation, 
We linearize Eq. \ref{nearly_iso_explicit} about the above
solution for $\beta\equiv0$ and write 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, \delta v_z\to \delta v_z+\delta
  v_{z1},\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align}
which implies 
\begin{align}
  \chi \to 1 + \delta\chi = 1 + \ii \hat{\sigma}\left(\gamma-1\right)\delta\beta,
\end{align}
with $\delta v_z$ and $\hat{\sigma}$ given by Eq. \ref{pert_basic}. We
may then seek  
\begin{align}
  \delta v_{z1} = a \zhat^2 + b,
\end{align}
where $a$, $b$ are constants.

@@

A finite thermal relaxation time introduces buoyancy
forces, which is stabilizing for sub-adiabatically stratified disks
($\gamma > 1$). The $\zhat^2$ dependence in $\delta v_{z1}$ then makes
sense because it becomes significant for large $|\zhat|$, i.e. away from
the midplane where the effect of buoyancy first appears as $\beta$ is
increased from zero.

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI mode for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
$\khat\gg1$, we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
This is a pair of simultaneous equations for
$\hat{\omega}$ and $\beta$. In Appendix \ref {iso_stable} we discuss
stable solutions without vertical shear which yield
$\hat{\omega}\propto \khat^{-1}$ for $\khat\gg 1$
(Eq. \ref{iso_stable_disp}).  We are thus motivated to seek solutions
to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% we wish to seek a criterion independent of k 
% \begin{align}
%   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
%\end{align}
\begin{align}
  \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
  |\hat{\omega}\khat|=O(1). 
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}\label{beta_crit0}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$.

From our discussion in \S\ref{relax_pert}, which indicate introducing
a finite thermal relaxation is stabilizing, we expect
Eq. \ref{beta_crit0} is an upper limit. For a thin  
disk $\epsilon\ll 1$, we thus infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. That is, when
$\beta<\beta_\mathrm{crit}$, perturbations are  effectively isothermal. 
We will confirm Eq. \ref{iso_vsi_cond} numerically in
\S\ref{bcrit_num_test}.
